---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE
  , echo = FALSE
  , comment = "#>"
  , fig.path = "README_files/README-"
  , out.width = "100%"
)
includeFigure = function(x) {
  knitr::include_graphics(file.path("README_files", x))
}
```

```{r}
includeFigure("RdistanceTopBanner.PNG")
```

<!-- badges: start -->
<center> 

[![CRAN Version](http://www.r-pkg.org/badges/version/Rdistance)](http://www.r-pkg.org/pkg/Rdistance)
[![CRAN Downloads](http://cranlogs.r-pkg.org/badges/Rdistance)](http://cran.rstudio.com/web/packages/Rdistance/index.html)
[![CRAN Total Downloads](http://cranlogs.r-pkg.org/badges/grand-total/Rdistance)](http://www.r-pkg.org/pkg/Rdistance)
[![codecov](https://codecov.io/gh/tmcd82070/Rdistance/branch/master/graph/badge.svg)](https://codecov.io/gh/tmcd82070/Rdistance)
[![Lifecycle: stable](https://img.shields.io/badge/lifecycle-stable-brightgreen.svg)](https://lifecycle.r-lib.org/articles/stages.html#stable)

[![](https://img.shields.io/github/commit-activity/m/tmcd82070/Rdistance)](https://github.com/tmcd82070/Rdistance/commits/master)
![](https://img.shields.io/github/commits-since/tmcd82070/Rdistance/v3.0.0/master?color=f9930d)
![](https://img.shields.io/github/contributors/tmcd82070/Rdistance?color=656d44)
![](https://img.shields.io/github/last-commit/tmcd82070/Rdistance?color=8e755e)
![](https://img.shields.io/github/release-date/tmcd82070/Rdistance?color=c95443)

</center> 

<!-- [![Travis-CI Build Status](https://travis-ci.org/tmcd82070/Rdistance.svg?branch=master)](https://travis-ci.org/tmcd82070/Rdistance) -->
<!-- badges: end -->

## Simplified Distance-Sampling in R

**Rdistance** analyzes line- and point-transect distance-sampling data. If you are unfamiliar with distance-sampling, check out our primer, [Distance Sampling for the Average Joe](https://github.com/tmcd82070/Rdistance/wiki/Distance-Sampling-for-the-Average-Joe). If you are ready to take on an analysis, the best place to start is the **Examples** section (below).  The *Example* sections of help files `?RdistDf`, `?dfuncEstim`, and `?abundEstim` contain enough information to get you started and explain a few details of analysis in **Rdistance**. The **Rdistance** [wiki](https://github.com/tmcd82070/Rdistance/wiki) (always a work-in-progress) contains additional information on distance sampling and is the equivalent of vignettes. 


```{r}
includeFigure("RdistanceSeparator.PNG")
```

## Features

* Line-transect and point-transects 
* Likelihood functions: 
  + half-normal (`halfnorm`) 
  + hazard rate (`hazrate`) 
  + negative exponential (`negexp`)
* Expansion terms
* Standard methods: `print`, `plot`, `predict`, `AIC`, etc.
* Observation and transect-level distance function covariates
* Standard R formula model specification (e.g., `distance ~ elevation + observer`)
* Measurement unit control and automatic conversion
* Automated bootstrap confidence intervals 
* Overall (study area) and site level abundance estimates
* Help and wiki content taught at multiple workshops and reviewed by multiple authors

    
```{r}
includeFigure("RdistanceSeparator.PNG")
```


## Installation


**Rdistance** depends on and imports a number of other packages. All installations require compilation because some dependent packages require compilation. Compilation requires either `Rtools` (on Windows systems) or `r-base-dev` (on Linux systems). The number of dependencies is large, in part because `Rdistance` imports `dplyr`, so a fresh installation will require patience.  


### Windows - `Rtools`: 

1.  Install `Rtools`: Download the correct `Rtools` installer for your version of R from [here](https://cran.r-project.org/bin/windows/Rtools/).  When download completes, run the installer (double-click it) and accept all default options. If `Rtools` is already installed, skip this step.


### Ubuntu - `r-base-dev`:

If R is installed, chances are repository lists are correct. For a fresh installation (of R) follow the instructions [here](https://cran.r-project.org/bin/linux/ubuntu/fullREADME.html).  This will insure that `apt get` has the correct list of repositories.

1.  Install the equivalent of `Rtools` with the following command:

```
sudo apt-get install r-base-dev
```

2.  `Rdistance` requires the `units` package, and the `units` package requires the `udunits` library. Install the `udunits` library on linux via:

```
sudo apt-get install libudunits2-dev
```

### Install Rdistance :

Once R is installed and running, install the current stable release or the development release with the following commands:

1.  ***Current release***: Install the current release directly from CRAN.  In the R terminal, issue...

``` r
install.packages("Rdistance")
```

2.  ***Development version***: `Rdistance` is under active development, and the 'development' version often contains patches between official releases.  Inspect commit messages for commits following the most recent release for a description of the patches. Install the development version from [GitHub](https://github.com/tmcd82070/Rdistance) using:

``` r
if( !require("devtools") ){
  install.packages("devtools")
}
devtools::install_github("tmcd82070/Rdistance")
```

```{r}
includeFigure("RdistanceSeparator.PNG")
```


## Tutorials 

Beginner, intermediate, and advanced `Rdistance` tutorials are [here](https://mcdonalddatasciences.com/Rdistance.html).


```{r}
includeFigure("RdistanceSeparator.PNG")
```


# RECENT CHANGES

Our [NEWS](NEWS.md) file contains salient changes across version numbers.


