---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE
  , echo = FALSE
  , comment = "#>"
  , fig.path = "README_files/README-"
  , out.width = "100%"
)
includeFigure = function(x) {
  knitr::include_graphics(file.path("README_files", x))
}
```

```{r}
includeFigure("RdistanceTopBanner.PNG")
```

<!-- badges: start -->
<center> 

[![CRAN Version](http://www.r-pkg.org/badges/version/Rdistance)](http://www.r-pkg.org/pkg/Rdistance)
[![CRAN Downloads](http://cranlogs.r-pkg.org/badges/Rdistance)](http://cran.rstudio.com/web/packages/Rdistance/index.html)
[![CRAN Total Downloads](http://cranlogs.r-pkg.org/badges/grand-total/Rdistance)](http://www.r-pkg.org/pkg/Rdistance)
[![codecov](https://codecov.io/gh/tmcd82070/Rdistance/branch/master/graph/badge.svg)](https://codecov.io/gh/tmcd82070/Rdistance)
[![Lifecycle: stable](https://img.shields.io/badge/lifecycle-stable-brightgreen.svg)](https://lifecycle.r-lib.org/articles/stages.html#stable)

[![](https://img.shields.io/github/commit-activity/m/tmcd82070/Rdistance)](https://github.com/tmcd82070/Rdistance/commits/master)
![](https://img.shields.io/github/commits-since/tmcd82070/Rdistance/v3.0.0/master?color=f9930d)
![](https://img.shields.io/github/contributors/tmcd82070/Rdistance?color=656d44)
![](https://img.shields.io/github/last-commit/tmcd82070/Rdistance?color=8e755e)
![](https://img.shields.io/github/release-date/tmcd82070/Rdistance?color=c95443)

</center> 

<!-- [![Travis-CI Build Status](https://travis-ci.org/tmcd82070/Rdistance.svg?branch=master)](https://travis-ci.org/tmcd82070/Rdistance) -->
<!-- badges: end -->

## Simplified Distance-Sampling in R

**Rdistance** analyzes line- and point-transect distance-sampling data. If you are unfamiliar with distance-sampling, check out our primer, [Distance Sampling for the Average Joe](https://github.com/tmcd82070/Rdistance/wiki/Distance-Sampling-for-the-Average-Joe). If you are ready to take on an analysis, the best place to start is the **Examples** section (below).  The *Example* sections of help files `?RdistDf`, `?dfuncEstim`, and `?abundEstim` contain enough information to get you started and explain a few details of analysis in **Rdistance**. The **Rdistance** [wiki](https://github.com/tmcd82070/Rdistance/wiki) (always a work-in-progress) contains additional information on distance sampling and is the equivalent of vignettes. 


```{r}
includeFigure("RdistanceSeparator.PNG")
```

## Features

* Line-transect and point-transects 
* Likelihood functions: 
  + half-normal (`halfnorm`) 
  + hazard rate (`hazrate`) 
  + negative exponential (`negexp`)
* Expansion terms
* Standard methods: `print`, `plot`, `predict`, `AIC`, etc.
* Observation and transect-level distance function covariates
* Standard R formula model specification (e.g., `distance ~ elevation + observer`)
* Measurement unit control and automatic conversion
* Automated bootstrap confidence intervals 
* Overall (study area) and site level abundance estimates
* Help and wiki content taught at multiple workshops and reviewed by multiple authors

    
```{r}
includeFigure("RdistanceSeparator.PNG")
```


## Installation

### Prerequisites

**Rdistance** depends on and imports a number of other packages. For a fresh installation, you will need to install `Rtools` prior to installing `Rdistance` because some dependent packages require compilation.  If `Rtools` is already installed, skip this step. 

***Windows***: Install `Rtools` by first downloading the correct installer for your version of R from [here](https://cran.r-project.org/bin/windows/Rtools/).  Once downloaded, run the installer (double-click it) and accept all default options.

***Linux***: Install the equivalent of `Rtools` using the following command:

```
sudo apt-get install r-base-dev
```

The `units` package will require the `udunits` library which can be installed on linux via:

```
sudo apt-get install libudunits2-dev
```

### Install Rdistance

The current release's code is [here](https://github.com/tmcd82070/Rdistance/releases). Install the current release in R directly from CRAN using:

``` r
install.packages("Rdistance")
```

This will check for and install all dependencies.  The number of dependencies is large, in part because `Rdistance` imports `dplyr`, so a fresh installation will require patience.  

**Rdistance** is under active development.  Install the development version from [GitHub](https://github.com/) using:
``` r
if( !require("devtools") ){
  install.packages("devtools")
}
devtools::install_github("tmcd82070/Rdistance")
```


```{r}
includeFigure("RdistanceSeparator.PNG")
```

## Examples 


These examples show basic estimation of abundance via distance-sampling analyses, both with and without covariates. 

### Line Transects - No Covariates

```{r sparrowData, echo=TRUE}
library(Rdistance)

# Example data
data("sparrowDetectionData")  # access example data
data("sparrowSiteData")
head(sparrowDetectionData)  # inspect data
head(sparrowSiteData)

# Make nested data frame required by Rdistance
sparrowDf <- RdistDf(sparrowSiteData
                   , sparrowDetectionData
                   , by = "siteID"
                   , pointSurvey = FALSE
                   , observer = "single"
                   , .detectionCol = "detections"
                   , .effortCol = "length")
head(sparrowDf)
summary(sparrowDf, formula = dist ~ groupsize(groupsize))
# see detections on transect A1
sparrowDf |> 
  dplyr::filter(siteID == "A1") |>
  dplyr::reframe(detections)
# see detections on 5th transect
sparrowDf$detections[[5]]
```

#### Estimate hazard rate distance function

```{r lineHazrateExample, echo=TRUE}
# Set upper (right) truncation distance
whi <- set_units(200, "m")
# Fit hazard rate likelihood and estimate density per ha
oneHectare <- set_units(1, "ha")

# To save computation time, only compute point estimates.
# change `ci = NULL` to `ci = 0.95` to compute confidence intervals
dfuncFit <- sparrowDf |>
  dfuncEstim(dist ~ groupsize(groupsize)
           , likelihood = "hazrate"
           , w.hi = whi) |> 
  abundEstim(area = oneHectare
           , ci = NULL)
summary(dfuncFit)
plot(dfuncFit, nbins = 30, border = NA)

# Estimate stored in output object
dfuncFit$estimates
```


### Line Transects - With Vegetation Covariate

```{r lineHazrateExampleCovars, echo=TRUE}
# This time, estimates only. No confidence intervals
dfuncFit <- sparrowDf |>
  dfuncEstim(dist ~ bare + groupsize(groupsize)
           , likelihood = "hazrate"
           , w.hi = whi) |> 
  abundEstim(area = oneHectare
           , ci = NULL)

summary(dfuncFit)
plot(dfuncFit
   , newdata = data.frame(bare = c(30, 40, 50))
   , lty = 1
   , nbins = 30
   , border = NA
   , col = "grey75")
```

### Point Transects - No Covariates

```{r thrasherData, echo=TRUE}
# Example data
data("thrasherDetectionData")  # access example data
data("thrasherSiteData")
head(thrasherDetectionData)  # inspect example data
head(thrasherSiteData)

# Make nested data frame required by Rdistance
thrasherDf <- RdistDf(thrasherSiteData
                   , thrasherDetectionData
                   , by = "siteID"
                   , pointSurvey = TRUE
                   , observer = "single"
                   , .detectionCol = "detections"
                   , .effortCol = "npoints")
head(thrasherDf)
summary(thrasherDf, formula = dist ~ groupsize(groupsize))
```

```{r pointHazrateExample, echo=TRUE}
dfuncFit <- thrasherDf |>
  dfuncEstim(dist ~ groupsize(groupsize)
           , likelihood = "hazrate") |> 
  abundEstim(area = oneHectare
           , ci = NULL)

summary(dfuncFit)
plot(dfuncFit
   , nbins = 15)
```

### Point Transects - With Vegetation Covariates

```{r pointHazrateExampleCovars, echo=TRUE}
dfuncFit <- thrasherDf |>
  dfuncEstim(dist ~ bare + shrub + groupsize(groupsize)
           , likelihood = "hazrate") |> 
  abundEstim(area = oneHectare
           , ci = NULL)
summary(dfuncFit)
plot(dfuncFit
   , newdata = data.frame(bare = c(30, 35, 40)
                       , shrub = 20)
   , lty = 1)
```

<!-- github CI examples for auto rendering of Rmd <https://github.com/r-lib/actions/tree/v1/examples>. -->

```{r}
includeFigure("RdistanceSeparator.PNG")
```


# RECENT CHANGES

See our [NEWS](NEWS.md) file for changes across version numbers.


