---
title: "Extended `dfuncEstim` examples"
author: "Trent McDonald"
date: "`r format(Sys.Date(), '%F')`"
output: pdf_document
vignette: >
  %\VignetteIndexEntry{Rdistance: Extended dfuncEstim Examples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteKeyword{ds,distance,intro,transect,abundance,density,wildlife}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Following is a series of `dfuncEstim` calls showing calling parameters for many popular distance sampling options. 

```{r}
library(Rdistance)
data("sparrowDetectionData")
data("sparrowSiteData")
```

# Half-normal function with right truncation

```{r}
dfunc <- dfuncEstim(formula = dist ~ 1
                  , detectionData = sparrowDetectionData
                  , w.hi = units::set_units(100, "m"))
dfunc
plot(dfunc, col="grey")
```

# Half-normal function, truncation, group sizes

```{r}
dfunc <- dfuncEstim(formula = dist ~ groupsize(groupsize)
                  , detectionData = sparrowDetectionData
                  , w.hi = units::set_units(100, "m"))
dfunc
plot(dfunc, col="grey")
```

# Half-normal function with factor covariate 

Increase the maximum number of iterations if distance function is having trouble converging.  Here, the `observer` covariate is in the *site* data frame (`sparrowSiteData`), so this data frame must be included in the call to `dfuncEstim`. 

```{r}
dfuncObs <- dfuncEstim(formula = dist ~ observer
                     , detectionData = sparrowDetectionData
                     , siteData = sparrowSiteData
                     , w.hi = units::set_units(100, "m")
                     , control=RdistanceControls(maxIter=1000))
dfunc
plot(dfunc, col="grey")
```

# Hazard-rate function with covariate, truncation, and variable group sizes

Group sizes are not used to estimate the distance function.  Group sizes are used in future calls to `abundEstim`. 

```{r}
dfuncObs <- dfuncEstim(formula = dist ~ observer + groupsize(groupsize)
                     , likelihood = "hazrate"
                     , detectionData = sparrowDetectionData
                     , siteData = sparrowSiteData
                     , w.hi = units::set_units(100, "m"))
dfunc
plot(dfunc, col="grey")
```

# Hazard rate, covariate, group sizes, left and right truncation

```{r}
dfunc <- dfuncEstim(formula = dist ~ observer + groupsize(groupsize)
                     , likelihood = "hazrate"
                     , detectionData = sparrowDetectionData
                     , siteData = sparrowSiteData
                     , w.lo = units::set_units(20, "m")
                     , x.scl = units::set_units(20, "m")
                     , w.hi = units::set_units(100, "m"))
dfunc
plot(dfunc, col="grey")
```



# Hazard rate, covariate, group sizes, specify g(0)

Specify $g(0)$ at the intercept by setting `g.x.scl`.  This scales the entire distance function. 

```{r}
dfunc <- dfuncEstim(formula = dist ~ observer + groupsize(groupsize)
                     , likelihood = "hazrate"
                     , detectionData = sparrowDetectionData
                     , siteData = sparrowSiteData
                     , g.x.scl = 0.8)
dfunc
plot(dfunc, col="grey")
```

