---
title: "Modeling Abundance in Relation to Covariates"
author: "Jason D. Carlisle and Trent L. McDonald"
date: "July 20, 2017"
output: pdf_document
vignette: >
  %\VignetteIndexEntry{Rdistance_CovariatesOnAbundance}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

# Introduction
This tutorial builds on the introductory material presented in the **Rdistance Tutorial for Beginners**.  The goal of a distance-sampling study is often to relate abundance to some other variable, e.g., a covariate describing the habitat at each survey site.  Hierarchical distance sampling methods have been developed which combine the estimation of detection probability and abundance in one modeling framework (Royle et al. 2004).  But traditionally, a two-staged approach (?? Buckland et al. 2009 ??) has been used to relate abundance to covariates when using distance sampling:  first, model the detection function and inflate the observed counts based on the estimated detection probability to estimate density at each sample unit (e.g., transect); second, regress density on covariates.  Following such an approach, `Rdistance` can provide you with site-level estimates of abundance that are corrected for imperfect detection (stage 1). You can then use these estimates as the response variable in one of the many modeling frameworks available in Program R that you're likely already familiar with (e.g., `lm`, `glm`, `lme4`, etc.; stage 2).  Here, we make use of the example sparrow dataset contained within `Rdistance` (i.e., line transect surveys of Brewer's Sparrows), to demonstrate an example workflow of a conventional two-stage analytical approach.  This tutorial is current as of version `r packageVersion("Rdistance")` of `Rdistance`.

# Estimate abundance at each site
The process of generating an overall, study-area estimate of abundance is outlined in the **Rdistance Tutorial for Beginners**.  In contrast, here we want an abundance estimate specific to each site, not the study area as a whole.  We can obtain an abundance estimate for each site surveyed using the `by.id` argument (`by.id=TRUE`) within `F.abund.estim` or `F.automated.CDA`.

```{r}
require(Rdistance)
data(sparrow.detections)
data(sparrow.transects)
auto <- F.automated.CDA(detection.data=sparrow.detections,
                        transect.data=sparrow.transects,
                        by.id=TRUE, w.hi=150, area=10000, R=10, ci=0.95,
                        plot=FALSE, plot.bs=FALSE)

```

Because `by.id=TRUE`, a data.frame of the site-level abundances is stored in the `auto` object under the name `nhat.df`:

```{r}
head(auto$nhat.df)
```

Because distances were measured in meters and `area=10000`, abundance (density) estimates are given as the number of sparrows per hectare (ha).

# Model abundance estimates

The original transect dataset contained a variable named `sagemean` that represents the average sagebrush cover at each transect.  Suppose we want to examine the relationship between sagebrush cover and sparrow abundance.  First, merge the new data.frame of site-level abundance estimates back to the orignial transect data.frame that includes `sagemean`.


```{r}
mydata <- merge(sparrow.transects, auto$nhat.df, by="siteID")
head(mydata)
```


We now have both variables in the same data.frame.  For the sake of demonstration, we will fit a simple linear regression model using the `lm` function to assess the relationship between sagebrush cover and sparrow density (per ha).  We see that sparrow density increases with sagebrush cover, and density nearly quadruples across the range of sagebrush cover surveyed.

```{r}
mod <- lm(nhat ~ sagemean, data=mydata)
summary(mod)
```

```{r, fig.width=4.5, fig.height=5, echo=FALSE}
plot(mydata$nhat ~ mydata$sagemean,
     xlab="Sagebrush Cover (%)", ylab="Sparrow Density (birds per ha +/- 95% CI)")

pred.sage <- seq(min(mydata$sagemean), max(mydata$sagemean), by=0.5)
pred.nhat <- data.frame(predict(mod, newdata=data.frame(sagemean=pred.sage),
                                interval="confidence"))

lines(pred.sage, pred.nhat$fit, col="red", lwd=3)
lines(pred.sage, pred.nhat$lwr, col="red", lwd=2, lty="dashed")
lines(pred.sage, pred.nhat$upr, col="red", lwd=2, lty="dashed")

```


# Conclusion
By using `Rdistance` to estimate site-level abundance, you can account for imperfect detection, and use the modeling framework you choose to then relate site-level covariates to abundance.  We demonstrated this workflow using `Rdistance`'s `F.automated.CDA` function to fit a detection function and provide abundance estimates for each site, and the standard `lm` function to model the relationship between a habitat covariate and sparrow density.